---
title: 'BRFSS R Package'
engine: knitr
type: slides
format:
  revealjs:
    transition: slide
    background-transition: fade
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---

```{r, include=FALSE}
library(tidyverse)
```

# Outline

- Dataset
  - Demographics
  - Cognitive Decline
  - Tobacco Use
- Cleaning the Data
  - Investigating Missing Values
  - Imputing Missing Values
  - Tidying the Dataset
- R Package
  - Example Usage
- Questions

# Dataset

## Demographics

## Cognitive Decline

## Tobacco Use

# Cleaning the Data

## Investigating Missing Values

## Imputing Missing Values

## Tidying the Dataset

# R Package

## Example Usage

To install and load the package:

```{r, results='hide', message=FALSE, warning=FALSE}
remotes::install_github("mfehr7/BRFSSMissouri")
library(BRFSSMissouri)
```

The datasets are automatically loaded in as global variables:

```{r, eval=FALSE}
# Complete dataset
missouri_brfss

# Sub-datasets
demographics
tobacco_use
cognitive_decline
```

As an example analysis, this code looks at the average `SMOKE100` value
(using `SMOKE100_IMP1`) among different ages.

```{r}
# Summary table
smoke_by_age <- missouri_brfss %>%
  group_by(AGE80) %>%
  summarise(`AVG_SMOKE100` = mean(SMOKE100_IMP1))

# Summary plot
smoke_by_age %>%
  ggplot(aes(x = AGE80, y = AVG_SMOKE100)) + geom_bar(stat = "identity")
```

A value of 1 indicates that a person *has* smoked 100 cigarettes, while
a value of 2 indicates that a person *has not*. The trend seen in the
plot is what you would expect, as cigarettes were more common for older
generations growing up than they are now. Younger ages have an average
closer to 2, since their generations have alternate, popular nicotine
devices.

# Questions
